Synthesizing Symmetric Lenses

### 2019-02-28 papers/SymmetricLenses/main.tex
More comments and polish

Even for small examples, lenses like the one above take a little getting used
to. For larger and more complex formats, it can become quite difficult to
write lenses by hand.  Instead, we wish to synthesize them from examples. 
That is, given a pair of regular expressions and a set of input-output examples, we want
to find a simple symmetric lens that is typed by the regular expression pair and
satisfies all the input/output examples. We call such a lens a \emph{satisfying
  lens}. In our running example, we wish to synthesize a lens between
\lstinline{emp_salaries} and \lstinline{emp_insurance}, using as an input-output
example the data in Figure~\ref{fig:minimized-representations}.

One challenge is that the simple symmetric lens combinators permit many
well-typed lenses
between a given pair of regular expressions. For example, 
Figure~\ref{fig:example_lens} gives one possible lens between regular
expressions  \lstinline{emp_salaries} and \lstinline{emp_insurance},
but 
%
\begin{lstlisting}
disconnect(emp_salaries, emp_insurance, ..., ...)
\end{lstlisting}
%
is another well-typed lens that satisfies the example in
Figure~\ref{fig:minimized-representations} (where \lstinline{...} represents
components of the example itself). In general, {\em many} examples may be
required to rule out all possible occurrences of nested \Disconnect lenses,
particularly in complex formats.
%\ksf{I don't understand the following.  We shouldn't explain the issue in
%terms of \PutR and \PutL as we are focused on the combinators rather
%than the raw functions.}
%The default strings given to \Disconnect merely need to be the strings
%in the input-output examples, and the lens will still satisfy the
%specification.  While the disconnect lens is prohibited by two
%examples, or by a carefully chosen input-output example on \PutR
%or \PutL, this problem continues presenting itself for the sublenses,
%and is highlighted even more on larger formats.
While we could ask programmers to provide detailed and precise specifications
for their synthesis tasks, we would prefer an algorithm that can intelligently
choose a lens despite the underspecification. Instead of merely finding
\emph{any} satisfying lens, we wish to synthesize a satisfying lens that is
likely to please the user.

How can we identify such a ``likely'' lens? We propose the following heuristic:
A satisfying lens is ``more likely'' if it uses more data from one format to
construct the other. One manifestation of this principle involves choosing the
identity lens (which uses all the data) rather than the disconnect lens (which
uses none of the data). We define the likelihood of a satisfying lens as the
expected number of bits required to recover data in one format from synchronized
data in the other. Higher likelihoods correspond to fewer bits. Two
strings $s$ and $t$ are \emph{synchronized according to lens $l$} if $l.\PutR\
s\ t = t$ and $l.\PutL\ t\ s = s$. We can \emph{recover} $s$ from
$t$ using bits $b$ and lens $l$ if we can reconstruct $s$ from $t$, $b$, and
$l$. For example, given the \IdentityLens{} lens, we can recover $s$ from $t$
using no bits because, in this case, $s$ is just $t$. In contrast, given the
\Disconnect{} lens, we need enough bits to fully encode $s$ in order to recover
it from $t$ because $t$ contains no useful information. Thus, if both
\IdentityLens{} and \Disconnect{} are satisfying lenses for a particular pair of
formats, then \IdentityLens{} will be more likely.

>>>>

For larger and more complex formats, it can become quite difficult to
write lenses by hand, as we did above.  Instead, we would like to synthesize
them from types and examples.  
That is, given a pair of regular expressions and a set of input-output
examples, we want 
to find a simple symmetric lens that is typed by the regular expression pair and
satisfies all the input/output examples. We this a \emph{satisfying
  lens}. For our running example, suppose we wish to synthesize a lens between
\lstinline{emp_salaries} and \lstinline{emp_insurance}, using as sole
input-output example the data in Figure~\ref{fig:minimized-representations}.

One challenge is that the simple symmetric lens combinators permit many
well-typed lenses
between a given pair of regular expressions. For example, 
Figure~\ref{fig:example_lens} gives one possible lens between regular
expressions  \lstinline{emp_salaries} and \lstinline{emp_insurance},
but 
%
\begin{lstlisting}
disconnect(emp_salaries, emp_insurance, ..., ...)
\end{lstlisting}
%
is another well-typed lens that satisfies the example in
Figure~\ref{fig:minimized-representations} (where \lstinline{...} represents
components of the example itself\bcp{huh?}). In general, {\em many} examples
may be 
required to rule out all possible occurrences of nested\bcp{??} \Disconnect lenses,
particularly in complex formats.
%\ksf{I don't understand the following.  We shouldn't explain the issue in
%terms of \PutR and \PutL as we are focused on the combinators rather
%than the raw functions.}
%The default strings given to \Disconnect merely need to be the strings
%in the input-output examples, and the lens will still satisfy the
%specification.  While the disconnect lens is prohibited by two
%examples, or by a carefully chosen input-output example on \PutR
%or \PutL, this problem continues presenting itself for the sublenses,
%and is highlighted even more on larger formats.
\proposedelete{While we could ask programmers to provide detailed and precise specifications
for their synthesis tasks, we would prefer an algorithm that can intelligently
choose a lens despite the underspecification.} Instead of merely finding
\emph{any} satisfying lens, we wish to synthesize a satisfying lens that is
likely to please the user.

How can we identify such a ``likely'' lens? We propose the following heuristic:
A satisfying lens is ``more likely'' if it uses more data from one format to
construct the other.  For example, the
identity lens (which uses all the data) is more likely than the disconnect
lens (which uses none). Formally, we define the {\em likelihood} of a
satisfying lens as the 
expected number of bits required to recover data in one format from 
data in the other; higher likelihoods correspond to fewer bits. Two
strings $s$ and $t$ are \emph{synchronized according to lens
  $l$}\bcp{Earlier we used $\Lens$, now we are switching to $l$...} if $l.\PutR\
s\ t = t$ and $l.\PutL\ t\ s = s$. We can \emph{recover} $s$ from 
$t$ using bits $b$ and lens $l$ if we can reconstruct $s$ from $t$, $b$, and
$l$. For example, given the \IdentityLens{} lens, we can recover $s$ from $t$
using no bits because, in this case, $s$ is just $t$. In contrast, given the
\Disconnect{} lens, we need enough bits to fully encode $s$ in order to recover
it from $t$ because all the information in $t$ gets thrown away. Thus, if both
\IdentityLens{} and \Disconnect{} are satisfying lenses for a particular pair of
formats, \IdentityLens{} will be more likely.
